Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
minor fixen

Here we provide a concrete implementation of the simplification algorithm.
The algorithm uses a few simple structures
besides the node and edge tables discussed earlier. As the algorithm needs to track the mapping
between input and output node IDs over specific genomic intervals, we use a $\ancsegment$
type to track these individual mappings. Thus, a segment $x$ has three attributes:
$\attrleft{x}$ and $\attrright{x}$ represent the genomic interval in the usual way,
and $\attrnode{x}$ represents the output node ID that is assigned to this interval.
These output node IDs correspond to colors in the ``paint pot'' analogy.
The per-interval segments are assigned to input IDs via the $\Al$ mapping,
which is the main state required by the algorithm. Each element $\Al[u]$
is a list of ancestral segments, describing the output node IDs mapped to input
node ID $u$ over a set of intervals.
For each input node, we need to process the ancestral segments
to find overlaps (and hence coalescences), and to update the $\Al$.
We do this using a $\priorityq{}$ type, which maintains a list of ancestral
segments sorted by left coordinate. For a priority queue $Q$, the operation
$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}$
removes and returns the segment in $Q$ with the smallest left coordinate.
To obtain the segment currently in the queue with the smallest left value we
write $Q[0]$.

@@

\algstep{S0.}{Initialisation.}{
    Set $\Al_u \leftarrow \listnew{}$ for $0 \leq u < |\Nt_I|$. Then,
    for each $u$ in $S$, set $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ and
    call $\listappend{\Al[u]}{\ancsegment{0, L, v}}$. Then set $Q \leftarrow \priorityq{}$
    and $u \leftarrow 0$.
}

\algstep{S1.}{Select parent edges.}{
    Set $E \leftarrow \{e \in \Et_I : \attrparent{e} = u \}$.
    Set $v \leftarrow -1$.  If $|E| = 0$, go to \algref{S8}.
}

\algstep{S2.}{Insert edge/ancestry intersections.}{
    For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
    if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
    set $\ell \leftarrow \max(\attrleft{x}, \attrleft{e})$,
    $r \leftarrow \min(\attrright{x}, \attrright{e})$,
    $y \leftarrow \ancsegment{\ell, r, \attrnode{x}}$ and call
    $\pqpush{Q}{y}$.
}

\algstep{S3.}{Find segments with minimum left coordinate.}{
    Set $\ell \leftarrow \attrleft{Q[0]}$, $r \leftarrow L$ and $X \leftarrow \mbox{List}()$.
    Then, while $|Q| > 0$ and $\attrleft{Q[0]} = \ell$, set $x \leftarrow \pqpop{Q}$,
    $r \leftarrow \min(r, \attrright{x})$ and call $\listappend{X}{x}$.
    Afterwards, if $|Q| > 0$, set $r \leftarrow \min(r, \attrleft{Q[0]})$. If
    $|X| > 1$ go to \algref{S5}.
}

\algstep{S4.}{No overlapping segments.}{
    Set $x \leftarrow X[0]$ and $\alpha \leftarrow x$. If $|Q| > 0$ and
    $\attrleft{Q[0]} < \attrright{x}$, set $\alpha \leftarrow
    \ancsegment{\attrleft{x}, \attrright{Q[0]}, \attrnode{x}}$,
    $\attrleft{x} \leftarrow \attrleft{Q[0]}$ and call $\pqpush{Q}{x}$.
    Go to \algref{S7}.
}

\algstep{S5.}{Overlap; new output node.}{
    If $v = -1$, set $v \leftarrow \taddrow{\Nt_0}{\Nt_I[u]}$.
}

\algstep{S6.}{Record edges.}{
    Set $\alpha \leftarrow \ancsegment{\ell, r, v}$. Then for each $x \in X$
    call $\taddrow{\Et_O}{l, r, v, \attrnode{x}}$ and, if $\attrright{x} > r$,
    set $\attrleft{x} \leftarrow r$ and call $\pqpush{Q}{x}$.
}

\algstep{S7.}{Left coordinate loop.}{
    Call $\listappend{\Al[u]}{\alpha}$. If $|Q| > 0$ return to \algref{S3}.
}

\algstep{S8.}{Parent loop.}{
    Set $u \leftarrow u + 1$. If $u < |\Nt_I|$, go to \algref{S1}.
}

@@

The algorithm begins in \algref{S0} by allocating an empty list to store ancestral
segments for each of the $|\Nt_I|$ input nodes, and then creating the initial
state for the samples in $S$. For each input sample with node ID $u$, we
add a row in the output node table $\Nt_O$, and create a mapping for this
new node $v$. After this initial state has been created, we then allocate our
priority queue $Q$, and set the current input ID $u$ to $0$.
Recall that the function $\taddrow{\Nt}{t}$ adds a new node to the node table $\Nt$
with birth time $t$, and returns the ID (i.e., index) of that new node in $\Nt$.
So, the operation $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ looks up the birth time
for input node ID $u$ in $\Nt_I$, adds a new node with this time to $\Nt_O$,
and stores the resulting (output) node ID in $v$.

To step back up through the pedigree, steps \algref{S1}--\algref{S8}
consider each node $u$ in turn (note we are assuming that nodes are ordered by time-since-birth).
In \algref{S1} we first find all edges in the input where the parent is equal to
our current input ID $u$, and then set the corresponding output ID $v$ to $-1$.
Then, in \algref{S2} we find all intersections between these input edges and the
mapped ancestry segments of their child nodes.
For each of these intersections we
create a new segment $y$ and insert it into the priority queue.
(In the paint pot analogy,
this stores in $Q$ all segments of color that must be transferred to the ancestor
with input ID $u$.)

After finding all overlaps and filling the priority queue with the resulting ancestry
segments, we then need to process these segments. Steps \algref{S3}--\algref{S7}
loop over the segments in the queue, considering each in turn until the queue is empty.
Step \algref{S3} builds a list $X$ of all segments covering the current left-most coordinate
(along with some bookkeeping to keep track of where our next
right coordinate $r$ is), removing them from $Q$. If there is only one element in $X$
this means that we have no overlapping ancestral segments at this coordinate and
we proceed to \algref{S4}.

In this case, only a single segment covers the leftmost region of the segments,
so there is no change in the mapping between input and output IDs,
and the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S7}).
(The color in this segment of the offspring can simply be transferred to the parent.)
However, if other segments in $Q$ intersect with $x$
(since $x$ is the current left-most segment, these have left coordinate less then $\attrright{x}$),
then we can only transfer the portion of $x$ up until it overlaps with other segments.
Therefore, we change the left coordinate
of $x$ to this coordinate and reinsert it into $Q$ where it will be processed again
later, effectively cutting off (and propagating) the leftmost segment of $x$.
After doing this, we proceed to \algref{S7}, update the ancestry mapping for
the input node $u$, and loop back to \algref{S3} if $Q$ is not empty.

On the other hand, if there is more than one segment in $X$, we know that an
overlap among the ancestral segments has occurred and we must update the output
nodes and edges accordingly. In step \algref{S5} we first allocate a new output
node $v$ for input node $u$, if it has not been done already (we may have many overlapping
segments along the genome). Step \algref{S6} then continues by creating a new
ancestry segment $\alpha$ mapping the current interval to the output node $v$.
We then iterate over all segments in $X$, adding the appropriate output edges
and editing and reinserting the ancestry segment $x$ into the priority queue if
necessary, as before.

>>>>

Here we provide a concrete implementation of the simplification algorithm.
The algorithm uses a few simple structures
besides the node and edge tables discussed earlier. As the algorithm needs to track the mapping
between input and output node IDs over specific genomic intervals, we use a $\ancsegment$
type to track these individual mappings. Thus, a segment $x$ has three attributes:
$\attrleft{x}$ and $\attrright{x}$ represent the genomic interval in the usual way,
and $\attrnode{x}$ represents the output node ID that is assigned to this interval.
These output node IDs correspond to colors in the ``paint pot'' analogy.
The per-interval segments are assigned to input IDs via the $\Al$ mapping,
which is the main state required by the algorithm. Each element $\Al[u]$
is a list of ancestral segments, describing the output node IDs mapped to input
node ID $u$ over a set of intervals.
For each input node, we need to process the ancestral segments
to find overlaps (and hence coalescences), and to update $\Al$.
We do this using a $\priorityq{}$ type, which maintains a list of ancestral
segments sorted by left coordinate. For a priority queue $Q$, the operation
$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}$
removes and returns the segment in $Q$ with the smallest left coordinate.
To obtain the segment currently in the queue with the smallest left value we
write $Q[0]$.

@@

\algstep{S1.}{Initialisation.}{
    Set $\Al_u \leftarrow \listnew{}$ for $0 \leq u < |\Nt_I|$. Then,
    for each $u$ in $S$, set $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ and
    call $\listappend{\Al[u]}{\ancsegment{0, L, v}}$. Then set $Q \leftarrow \priorityq{}$
    and $u \leftarrow 0$.
}

\algstep{S2.}{Select parent edges.}{
    Set $E \leftarrow \{e \in \Et_I : \attrparent{e} = u \}$.
    Set $v \leftarrow -1$.  If $|E| = 0$, go to \algref{S9}.
}

\algstep{S3.}{Insert edge/ancestry intersections.}{
    For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
    if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
    set $\ell \leftarrow \max(\attrleft{x}, \attrleft{e})$,
    $r \leftarrow \min(\attrright{x}, \attrright{e})$,
    $y \leftarrow \ancsegment{\ell, r, \attrnode{x}}$ and call
    $\pqpush{Q}{y}$.
}

\algstep{S4.}{Find segments with minimum left coordinate.}{
    Set $\ell \leftarrow \attrleft{Q[0]}$, $r \leftarrow L$ and $X \leftarrow \mbox{List}()$.
    Then, while $|Q| > 0$ and $\attrleft{Q[0]} = \ell$, set $x \leftarrow \pqpop{Q}$,
    $r \leftarrow \min(r, \attrright{x})$ and call $\listappend{X}{x}$.
    Afterwards, if $|Q| > 0$, set $r \leftarrow \min(r, \attrleft{Q[0]})$. If
    $|X| > 1$ go to \algref{S6}.
}

\algstep{S5.}{No overlapping segments.}{
    Set $x \leftarrow X[0]$ and $\alpha \leftarrow x$. If $|Q| > 0$ and
    $\attrleft{Q[0]} < \attrright{x}$, set $\alpha \leftarrow
    \ancsegment{\attrleft{x}, \attrright{Q[0]}, \attrnode{x}}$,
    $\attrleft{x} \leftarrow \attrleft{Q[0]}$ and call $\pqpush{Q}{x}$.
    Go to \algref{S8}.
}

\algstep{S6.}{Overlap; new output node.}{
    If $v = -1$, set $v \leftarrow \taddrow{\Nt_0}{\Nt_I[u]}$.
}

\algstep{S7.}{Record edges.}{
    Set $\alpha \leftarrow \ancsegment{\ell, r, v}$. Then for each $x \in X$
    call $\taddrow{\Et_O}{l, r, v, \attrnode{x}}$ and, if $\attrright{x} > r$,
    set $\attrleft{x} \leftarrow r$ and call $\pqpush{Q}{x}$.
}

\algstep{S8.}{Left coordinate loop.}{
    Call $\listappend{\Al[u]}{\alpha}$. If $|Q| > 0$ return to \algref{S4}.
}

\algstep{S9.}{Parent loop.}{
    Set $u \leftarrow u + 1$. If $u < |\Nt_I|$, go to \algref{S2}.
}

@@

The algorithm begins in \algref{S1} by allocating an empty list to store ancestral
segments for each of the $|\Nt_I|$ input nodes, and then creating the initial
state for the samples in $S$. For each input sample with node ID $u$, we
add a row in the output node table $\Nt_O$, and create a mapping for this
new node $v$. After this initial state has been created, we then allocate our
priority queue $Q$, and set the current input ID $u$ to $0$.
Recall that the function $\taddrow{\Nt}{t}$ adds a new node to the node table $\Nt$
with birth time $t$, and returns the ID (i.e., index) of that new node in $\Nt$.
So, the operation $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ looks up the birth time
for input node ID $u$ in $\Nt_I$, adds a new node with this time to $\Nt_O$,
and stores the resulting (output) node ID in $v$.

To process ancestral information moving back up through the pedigree, steps \algref{S2}--\algref{S9}
consider each node $u$ in turn (note we are assuming that nodes are ordered by time-since-birth).
In \algref{S2} we first find all edges in the input where the parent is equal to
our current input ID $u$, and then set the corresponding output ID $v$ to $-1$.
Then, in \algref{S3} we find all intersections between these input edges and the
mapped ancestry segments of their child nodes.
For each of these intersections we
create a new segment $y$ and insert it into the priority queue.
(In the paint pot analogy,
this stores in $Q$ all segments of color that must be transferred to the ancestor
with input ID $u$.)

After finding all overlaps and filling the priority queue with the resulting ancestry
segments, we then need to process these segments. Steps \algref{S4}--\algref{S8}
loop over the segments in the queue, considering each in turn until the queue is empty.
Step \algref{S4} builds a list $X$ of all segments covering the current left-most coordinate
(along with some bookkeeping to keep track of where our next
right coordinate $r$ is), removing them from $Q$. If there is only one element in $X$
this means that we have no overlapping ancestral segments at this coordinate and
we proceed to \algref{S5}.

In this case, only a single segment covers the leftmost region of the segments,
so there is no change in the mapping between input and output IDs,
and the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S8}).
(The color in this segment of the offspring can simply be transferred to the parent.)
However, if other segments in $Q$ intersect with $x$
(since $x$ is the current left-most segment, these have left coordinate less then $\attrright{x}$),
then we can only transfer the portion of $x$ up until it overlaps with other segments.
Therefore, we change the left coordinate
of $x$ to this coordinate and reinsert it into $Q$ where it will be processed again
later, effectively cutting off (and propagating) the leftmost segment of $x$.
After doing this, we proceed to \algref{S8}, update the ancestry mapping for
the input node $u$, and loop back to \algref{S4} if $Q$ is not empty.

On the other hand, if there is more than one segment in $X$, we know that an
overlap among the ancestral segments has occurred and we must update the output
nodes and edges accordingly. In step \algref{S6} we first allocate a new output
node $v$ for input node $u$, if it has not been done already (we may have many overlapping
segments along the genome). Step \algref{S7} then continues by creating a new
ancestry segment $\alpha$ mapping the current interval to the output node $v$.
We then iterate over all segments in $X$, adding the appropriate output edges
and editing and reinserting the ancestry segment $x$ into the priority queue if
necessary, as before.
